---
title: "Narrative Abilities of Hearing-Impaired Children"
output:
  html_document: default
  pdf_document: default
date: "2023-06-21"
---


# Installing the required libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `psych` package is used to to gather statistics about the initial data with the `describe` function.
The `corrplot` package is used to create diagrams of correlation matrices.

```{r, result='hide', message=FALSE}
if(!require(psych)) install.packages("psych")
if(!require(corrplot)) install.packages("corrplot")
library(psych)
library(corrplot)
```

# Loading and preparing the data
The data file is imported and some initial transformations are applied to combine several `BRIEF` columns into one.
The data is then split up into the data for HI-children and the data for TD-children to use in separate analyses.

```{r}
data <- read.csv("./data.csv", stringsAsFactors=TRUE)

# Combine BRIEF_young_B and BRIEF_old_B to create a BRIEF_B column
data$BRIEF_B <- data$BRIEF_young_B
data$BRIEF_B[!is.na(data$BRIEF_old_B)] <- data$BRIEF_old_B[!is.na(data$BRIEF_old_B)]

# Combine BRIEF_young_C and BRIEF_old_C to create a BRIEF_C column
data$BRIEF_C <- data$BRIEF_young_C
data$BRIEF_C[!is.na(data$BRIEF_old_C)] <- data$BRIEF_old_C[!is.na(data$BRIEF_old_C)]

# Split the data between HI and TD:
data_HI <- data[data$HI == 1,]
data_TD <- data[data$HI != 1,]
```

## Descriptive statistics for the original data

The descriptive statistics of the whole dataset and the dataset splitted between HI- and TD-children are computed.

```{r, warning=FALSE}
descriprive_columns <- c("min", "max", "mean", "sd")

numeric_variables <- c(
  "age",
  "storylength_bus", 
  "MLU_bus",
  "mazes_bus", 
  "ungrammatical_bus", 
  "complex_bus", 
  "implicit_referents_bus",
  "plot_bus",
  "storylength_frog", 
  "MLU_frog", 
  "mazes_frog", 
  "ungrammatical_frog", 
  "complex_frog", 
  "implicit_referents_frog",
  "plot_frog",
  "BRIEF_B",
  "BRIEF_C",
  "NWRS",
  "hearing_loss_L", 
  "hearing_loss_R", 
  "speech_understanding_L", 
  "speech_understanding_R", 
  "speech_understanding_total",
  "age_rehabilitation"
)

describe(data)[numeric_variables, descriprive_columns]

numeric_variables_HI <- c(
  "age",
  "storylength_bus", 
  "MLU_bus", 
  "mazes_bus", 
  "ungrammatical_bus", 
  "complex_bus", 
  "implicit_referents_bus",
  "plot_bus",
  "storylength_frog", 
  "MLU_frog", 
  "mazes_frog", 
  "ungrammatical_frog", 
  "complex_frog", 
  "implicit_referents_frog",
  "plot_frog",
  "BRIEF_B",
  "BRIEF_C",
  "NWRS",
  "hearing_loss_L", 
  "hearing_loss_R", 
  "speech_understanding_L", 
  "speech_understanding_R", 
  "speech_understanding_total",
  "age_rehabilitation"
)
describe(data_HI)[numeric_variables_HI, descriprive_columns]

numeric_variables_TD <- c(
  "age",
  "storylength_bus", 
  "MLU_bus", 
  "mazes_bus", 
  "ungrammatical_bus", 
  "complex_bus", 
  "implicit_referents_bus",
  "plot_bus", 
  "storylength_frog", 
  "MLU_frog",
  "mazes_frog", 
  "ungrammatical_frog", 
  "complex_frog", 
  "implicit_referents_frog",
  "plot_frog",
  "BRIEF_B",
  "BRIEF_C",
  "NWRS"
)
describe(data_TD)[numeric_variables_TD, descriprive_columns]
```

## Contrasts for the data

Set contrasts and check that they were set correctly.

```{r}
# Binary variable `HI`
data$HI <- as.factor(data$HI)
contrast <- cbind (c(0.5, -0.5))   # HI = -0.5, TD = +0.5
colnames (contrast) <- c("-HI+TD")
contrasts (data$HI) <- contrast
contrasts (data$HI)
```

## Transforming the data to prepare for the models

A helper function is added that transforms the variables with a `sqrt`, `log` or cube root,
whenever one of these operations results in a lower skewness for that variable.
The operation with the lowest skewness is used,
and the variable is left unchanged when the operations do not lower the skewness.

```{r}
transform_column <- function(data, column_name) {
  column <- data[,column_name]
  minimum <- min(column[!is.na(column)])
  
  # log only works for values at least 1, but the skew is independent of this transformation.
  if (minimum <= 1) {
    column <- column - minimum + 1
  }
  
  all_skews_per_transformation <- c(
    "current" = abs(describe(column)[["skew"]]),
    "log" = abs(describe(log(column))[["skew"]]),
    "sqrt" = abs(describe(sqrt(column))[["skew"]]),
    "cube" = abs(describe(column^(1/3))[["skew"]])
  )
  
  transformations <- c(
    "current" = identity,  # Don't change the data when the skew is already minimized
    "log" = log,
    "sqrt" = sqrt,
    "cube" =  function(data) data^(1/3)
  )
  
  text_current = paste("For ", column_name, "the skew was", all_skews_per_transformation[["current"]])
  text_new = paste(names(which.min(all_skews_per_transformation)), min(all_skews_per_transformation))
  print(paste(text_current, "and with", text_new))

  transformation_minimising_skew <- transformations[[which.min(all_skews_per_transformation)]]
  
  return(transformation_minimising_skew(column))
}

for (numeric_variable in numeric_variables) {
  new_column <- transform_column(data, numeric_variable)
  data[numeric_variable] <- new_column
}
```

After these transformations, the data is can be scaled to set the mean to zero.

```{r}
data[numeric_variables] <- scale(data[numeric_variables], scale=FALSE)
```


## Checking correlations for the morphosyntactic and semantic-pragmatic structures

The correlations including their *p*-values are shown.

```{r}
corr.test(data[c("MLU_bus",  "complex_bus", "ungrammatical_bus")])$r
corr.test(data[c("MLU_bus",  "complex_bus", "ungrammatical_bus")])$p
corr.test(data[c("implicit_referents_bus", "storylength_bus")])$r
corr.test(data[c("implicit_referents_bus", "storylength_bus")])$p
corr.test(data[c("MLU_frog",  "complex_frog", "ungrammatical_frog")])$r
corr.test(data[c("MLU_frog",  "complex_frog", "ungrammatical_frog")])$p
corr.test(data[c("implicit_referents_frog", "storylength_frog")])$r
corr.test(data[c("implicit_referents_frog", "storylength_frog")])$p
```

# The effect of HI on narrative abilities

In this section the models are fitted and the results are shown.
The subsections each contain several models for the various dependent variables.
The several dependent and independent variables are listed at the start of each section for completeness.

## The Frog Story
Dependent variables (microstructures):

- Morphosyntax: `MLU_frog`, `ungrammatical_frog`, `complex_frog`
- Semantic pragmatics: `implicit_referents_frog`, `storylength_frog`
- Communicative language: `mazes_frog` 

Dependent variable (macrostructures): `plot_frog`

Independent variable: `HI`

```{r}
dependent_variables_frog_micro_morphosyntax <- c("MLU_frog", "complex_frog", "ungrammatical_frog")
dependent_variables_frog_micro_semanticpragmatics <- c("storylength_frog", "implicit_referents_frog")
dependent_variables_frog_micro_communicativelanguage <- c("mazes_frog")
dependent_variables_frog_macro <- c("plot_frog")

independent_variable <- c("HI")

model_MLU_frog <- lm(MLU_frog ~., data = data[,c("MLU_frog", independent_variable)])
summary(model_MLU_frog)
confint(model_MLU_frog, level = 0.95)

model_complex_frog <- lm(complex_frog ~., data = data[,c("complex_frog", independent_variable)])
summary(model_complex_frog)
confint(model_complex_frog, level = 0.95)

model_ungrammatical_frog <- lm(ungrammatical_frog ~., data = data[,c("ungrammatical_frog", independent_variable)])
summary(model_ungrammatical_frog)
confint(model_ungrammatical_frog, level = 0.95)

model_storylength_frog <- lm(storylength_frog ~., data = data[,c("storylength_frog", independent_variable)])
summary(model_storylength_frog)
confint(model_storylength_frog, level = 0.95)

model_implicit_referents_frog <- lm(implicit_referents_frog ~., data = data[,c("implicit_referents_frog", independent_variable)])
summary(model_implicit_referents_frog)
confint(model_implicit_referents_frog, level = 0.95)

model_mazes_frog <- lm(mazes_frog ~., data = data[,c("mazes_frog", independent_variable)])
summary(model_mazes_frog)
confint(model_mazes_frog, level = 0.95)

model_plot_frog <- lm(plot_frog ~., data = data[,c("plot_frog", independent_variable)])
summary(model_plot_frog)
confint(model_plot_frog, level = 0.95)
```


## The Bus Story
Dependent variables (microstructures):

- Morphosyntax: `MLU_bus`, `ungrammatical_bus`, `complex_bus`
- Semantic pragmatics: `implicit_referents_bus`, `storylength_bus`
- Communicative language: `mazes_bus` 

Dependent variable (macrostructures): `plot_bus`

Independent variable: `HI`

```{r}
dependent_variables_bus_micro_morphosyntax <- c("MLU_bus", "complex_bus", "ungrammatical_bus")
dependent_variables_bus_micro_semanticpragmatics <- c("storylength_bus","implicit_referents_bus")
dependent_variables_bus_micro_communicativelanguage <- c("mazes_bus")
dependent_variables_bus_macro <- c("plot_bus")

model_MLU_bus <- lm(MLU_bus ~., data = data[,c("MLU_bus", independent_variable)])
summary(model_MLU_bus)
confint(model_MLU_bus, level = 0.95)

model_complex_bus <- lm(complex_bus ~., data = data[,c("complex_bus", independent_variable)])
summary(model_complex_bus)
confint(model_complex_bus, level = 0.95)

model_ungrammatical_bus <- lm(ungrammatical_bus ~., data = data[,c("ungrammatical_bus", independent_variable)])
summary(model_ungrammatical_bus)
confint(model_ungrammatical_bus, level = 0.95)

model_storylength_bus <- lm(storylength_bus ~., data = data[,c("storylength_bus", independent_variable)])
summary(model_storylength_bus)
confint(model_storylength_bus, level = 0.95)

model_implicit_referents_bus <- lm(implicit_referents_bus ~., data = data[,c("implicit_referents_bus", independent_variable)])
summary(model_implicit_referents_bus)
confint(model_implicit_referents_bus, level = 0.95)

model_mazes_bus <- lm(mazes_bus ~., data = data[,c("mazes_bus", independent_variable)])
summary(model_mazes_bus)
confint(model_mazes_bus, level = 0.95)

model_plot_bus <- lm(plot_bus ~., data = data[,c("plot_bus", independent_variable)])
summary(model_plot_bus)
confint(model_plot_bus, level = 0.95)
```

# Correlations between phonological processing, executive functions and narrative abilities

The correlations of several variables regarding phonological processing,
executive functions and narrative abilities are computed.

## The effect of HI on the BRIEF and the NWR-S

```{r}
model_BRIEF_B <- lm(BRIEF_B ~., data = data[,c("BRIEF_B", "HI")])
summary(model_BRIEF_B)
confint(model_BRIEF_B, level = 0.95)

model_BRIEF_C <- lm(BRIEF_C ~., data = data[,c("BRIEF_C", "HI")])
summary(model_BRIEF_C)
confint(model_BRIEF_C, level = 0.95)

model_NWRS <- lm(NWRS ~., data = data[,c("NWRS", "HI")])
summary(model_NWRS)
confint(model_NWRS, level = 0.95)
```


## Correlations of the variables for The Frog Story for the all children

A helper function is used to plot the correlation matrices in a directory.

```{r}
new_colname <- function(column_name) {
  title <- column_name

  names_plot <- read.csv(file="./descriptive_colnames.csv")
  title <- names_plot[names_plot["old"] == column_name]

  if (length(title > 0 )) {  # Update the title if a rename exists
    title <- title[[2]]
  }

  return(title)
}

make_corrplot <- function(all_data, file_name, path_name="corrplots", cex=0.7, ncex=0.7) {
  dir.create(path_name, showWarnings = FALSE)
  file_path <- paste("./", path_name,"/", file_name, ".png", sep="")
  
  colnames(all_data) <- unlist(lapply(colnames(all_data), new_colname))

  png(height=1800, width=1800, file=file_path)
  corr_matrix <- corr.test(all_data)
  fctr <- 4
  
  corrplot(
      corr_matrix$r, title=paste("The Correlation Matrix For", file_name), p.mat = t(corr_matrix$p), insig="blank",
      cex.main=fctr,
      bg="gray92",
      tl.cex=fctr*cex,
      cl.cex=fctr*0.8,
      type="upper",
      diag = FALSE,
      method="color",
      addgrid.col = "darkgray",
      number.cex = fctr*ncex,
      mar = c(fctr*2, fctr*2, fctr*3, fctr*2), # margins
      tl.col="black",
      tl.offset=fctr*0.4
    )$corrPos -> corr_positions
  text(corr_positions$x, corr_positions$y, round(corr_positions$corr, 2), cex=fctr*ncex)
  
  # Surpress the output of the dev.of
  sink("tmp.log")
  dev.off()
  sink()
  file_removed <- file.remove("tmp.log")
}
```

Dependent variables (microstructures):

- Morphosyntax: `MLU_frog`, `ungrammatical_frog`, `complex_frog`
- Semantic pragmatics: `implicit_referents_frog`, `storylength_frog`
- Communicative language: `mazes_frog` 

Dependent variable (macrostructures): `plot_frog`

Independent variables: `BRIEF_B`, `BRIEF_C`, `NWRS`

```{r}
dependent_variables_frog <- c(
  "MLU_frog", 
  "complex_frog",
  "ungrammatical_frog",
  "storylength_frog", 
  "implicit_referents_frog",
  "mazes_frog",
  "plot_frog"
)

independent_variables_cognition <- c("BRIEF_B", "BRIEF_C", "NWRS")

make_corrplot(data[c(dependent_variables_frog, independent_variables_cognition)], "The Frog Story, NWR-S & BRIEF (total)")
corr.test(data[c(dependent_variables_frog, independent_variables_cognition)])$p
```

## Correlations of the variables for The Frog Story for the HI-children
Dependent variables (microstructures):

- Morphosyntax: `MLU_frog`, `ungrammatical_frog`, `complex_frog`
- Semantic pragmatics: `implicit_referents_frog`, `storylength_frog`
- Communicative language: `mazes_frog` 

Dependent variable (macrostructures): `plot_frog`

Independent variables: `BRIEF_B`, `BRIEF_C`, `NWRS`

```{r}
make_corrplot(data_HI[c(dependent_variables_frog, independent_variables_cognition)], "The Frog Story, NWR-S & BRIEF (HI)", cex=0.7, ncex=0.7)
corr.test(data_HI[c(dependent_variables_frog, independent_variables_cognition)])$p
```


## Correlations of the variables for The Frog Story for the TD-children
Dependent variables (microstructures):

- Morphosyntax: `MLU_frog`, `ungrammatical_frog`, `complex_frog`
- Semantic pragmatics: `implicit_referents_frog`, `storylength_frog`
- Communicative language: `mazes_frog` 

Dependent variable (macrostructures): `plot_frog`

Independent variables: `BRIEF_B`, `BRIEF_C`, `NWRS`

```{r}
make_corrplot(data_TD[c(dependent_variables_frog, independent_variables_cognition)], "The Frog Story, NWR-S & BRIEF (TD)")
corr.test(data_TD[c(dependent_variables_frog, independent_variables_cognition)])$p
```

## Correlations of the variables for The Bus Story for the all children
Dependent variables (microstructures):

- Morphosyntax: `MLU_bus`, `ungrammatical_bus`, `complex_bus`
- Semantic pragmatics: `implicit_referents_bus`, `storylength_bus`
- Communicative language: `mazes_bus` 

Dependent variable (macrostructures): `plot_bus`

Independent variables: `BRIEF_B`, `BRIEF_C`, `NWRS`

```{r}
dependent_variables_bus <- c(
  "MLU_bus", 
  "complex_bus",
  "ungrammatical_bus",
  "storylength_bus", 
  "implicit_referents_bus",
  "mazes_bus",
  "plot_bus"
)

make_corrplot(data[c(dependent_variables_bus, independent_variables_cognition)], "The Bus Story, NWR-S & BRIEF (total)")
corr.test(data[c(dependent_variables_bus, independent_variables_cognition)])$p
```

## Correlations of the variables for The Bus Story for the HI-children
Dependent variables (microstructures):

- Morphosyntax: `MLU_bus`, `ungrammatical_bus`, `complex_bus`
- Semantic pragmatics: `implicit_referents_bus`, `storylength_bus`
- Communicative language: `mazes_bus` 

Dependent variable (macrostructures): `plot_bus`

Independent variables: `BRIEF_B`, `BRIEF_C`, `NWRS`

```{r}
make_corrplot(data_HI[c(dependent_variables_bus, independent_variables_cognition)], "The Bus Story, NWR-S & BRIEF (HI)")
corr.test(data_HI[c(dependent_variables_bus, independent_variables_cognition)])$p
```

## Correlations of the variables for The Bus Story for the TD-children

Dependent variables (microstructures):

- Morphosyntax: `MLU_bus`, `ungrammatical_bus`, `complex_bus`
- Semantic pragmatics: `implicit_referents_bus`, `storylength_bus`
- Communicative language: `mazes_bus` 

Dependent variable (macrostructures): `plot_bus`

Independent variables: `BRIEF_B`, `BRIEF_C`, `NWRS`

```{r}
make_corrplot(data_TD[c(dependent_variables_bus, independent_variables_cognition)], "The Bus Story, NWR-S & BRIEF (TD)", cex=0.7, ncex=0.7)
corr.test(data_TD[c(dependent_variables_bus, independent_variables_cognition)])$p
```


# Correlations between rehabilitation and narrative abilities in HI-children

## Correlations of the rehabilitation variables for The Frog Story for the HI-children
Dependent variables (microstructures):

- Morphosyntax: `MLU_frog`, `ungrammatical_frog`, `complex_frog`
- Semantic pragmatics: `implicit_referents_frog`, `storylength_frog`
- Communicative language: `mazes_frog` 

Dependent variable (macrostructures): `plot_frog`

Independent variables: `hearing_loss_best_ear`, `age_rehabilitation`, `speech_understanding_total`

```{r}
independent_variables_rehabilitation <- c("hearing_loss_best_ear", "age_rehabilitation", "speech_understanding_total")

make_corrplot(data_HI[c(dependent_variables_frog, independent_variables_rehabilitation)], "The Frog Story, Hearing Loss, Age at the start of Rehabilitation & Speech Understanding")
corr.test(data_HI[c(dependent_variables_frog, independent_variables_rehabilitation)])$p
```


## Correlations of the rehabilitation variables for The Bus Story for the HI-children
Dependent variables (microstructures):

- Morphosyntax: `MLU_bus`, `ungrammatical_bus`, `complex_bus`
- Semantic pragmatics: `implicit_referents_bus`, `storylength_bus`
- Communicative language: `mazes_bus` 

Dependent variable (macrostructures): `plot_bus`

Independent variables: `hearing_loss_best_ear`, `age_rehabilitation`, `speech_understanding_total`

```{r}
make_corrplot(data_HI[c(dependent_variables_bus, independent_variables_rehabilitation)], "The Bus Story, Hearing Loss, Age at the start of Rehabilitation & Speech Understanding")
corr.test(data_HI[c(dependent_variables_bus, independent_variables_rehabilitation)])$p
```